---
title: "R for Data Science"
output: html_notebook
---

This notebook attempts to navigate Hadley Wickham's [R for Data Science](http://r4ds.had.co.nz/index.html).
```{r message=FALSE}
library(nycflights13)
library(tidyverse)
```


### Chapter 3 - Data Visualization
Initial question: Do cars with big engines use more fuel than cars with small engines? What does the relationship between engine size and fuel efficiency look like? 

[GGPlot2 Cheatsheet](http://rstudio.com/cheatsheets)

```{r}
mpg %>% head
```
To characterize the variables a little better, `displ` measures engine displacement, in litres, `year` corresponds to the year the car was manufactured, `drv` corresponds to front-wheel drive, rear-wheel drive, or four-wheel drive, `cty` is city miles per gallon, `hwy` is highway miles per gallon, and `fl` is fuel type.

```{r}
ggplot(data=mpg) + 
    geom_point(mapping=aes(x=displ, y=hwy))
```
This graph illustrates that fuel efficiency drops with increasing engine displacement (size). This confirms our hypothesis.

This also goes to show some ggplot2 basics:

1. Start with a call to `ggplot(data=<...>)`
2. Add on layers using `geom` functions, which use different geometric objects to represent data. [Types of Geoms](http://sape.inf.usi.ch/quick-reference/ggplot2/geom)
3. Supply `aes`(thetic) mappings to these `geom` functions to put the layer (with `x` and `y` variables) onto the plot.

More specifically, you can use the following as a plot template, replacing bracketed (`<...>`) sections: 
```
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
You can also add additional dimensions to the aesthetic (in addition to `x` and `y`), using `color`, `size`, `alpha` (transparency), or `shape` (max 6 unique vals). These dimensions do not need to vary by a variable, however -- can set manually within geom_point. Finally, variables can be categorical (discrete categories) or continuous (gradient colors, for example)
```{r}
ggplot(data=mpg) + 
    geom_point(mapping=aes(x=displ, y=hwy, color=class), shape=2) # 2 -> Open triangle
```
As we can see, SUVs and Pickups tend to have the largest engines and worst mileage, while midsize cars and minivans have average engine sizes and average mileage, and compact and subcompact cars have the smallest engines and highest highway mileage. One thing to note is that 2 seater cars (sports cars) are outliers -- they have very large engines but also average mileage, likely because they weight less than the SUVs and pickups with comparably sized engines.

Another way to add dimensions is with facets. `facet_wrap` is useful for faceting on a single variable, and `facet_grid` is useful for faceting on 2 variables. Note that faceting on a sufficiently lengthy continuous variable yields facets only for the (min, max) of the range
```{r}
ggplot(data=mpg) + 
    geom_point(mapping=aes(x=displ, y=hwy)) +
    facet_wrap(facet= ~class)   # Class is categorical

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

You can also overlap geoms simply. All three plots map the same data, but are special cases:
```{r message=FALSE}
# Explicit local mappings
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

# Concise global mapping
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

# Concise global mapping w/ explicit local mappings
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping=aes(color=class)) + 
  geom_smooth()
```

You can also group up best fit lines by another variable and color it the same for both geoms, in addition to jittering to prevent overfitting.
```{r message=FALSE, warning=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = class)) + 
  geom_point(position="jitter") + 
  geom_smooth(se = FALSE)
```

Building cool bar charts is also possible with GGPlot. For example, we could make a normal colored bar chart or a stacked bar chart
```{r}
ggplot(data=diamonds) + 
    geom_bar(mapping=aes(x=cut, fill=cut))

ggplot(data=diamonds) + 
    geom_bar(mapping=aes(x=cut, fill=clarity))
```

With additional complexity, we have an updated template
```  
ggplot(data = <DATA>) +  
  <GEOM_FUNCTION>(  
     mapping = aes(<MAPPINGS>),  
     stat = <STAT>,   
     position = <POSITION>  
  ) +  
  <COORDINATE_FUNCTION> +  
  <FACET_FUNCTION>  
```

### Chapter 5 - Data Transformation
Motivation: data is messy and rarely comes to you the way you would like it to be. Understanding how to manipulate data that you have possession of gives you deeper access to insights you can glean from the data.

[Dplyr Cheatsheet](http://rstudio.com/cheatsheets)

```{r}
flights %>% head
```

To briefly describe each variable:  

- `year`, `month`, `day` correspond to the date of departure  
- `dep_time`, `arr_time` correspond to actual departure and arrival times  
- `dep_delay`, `arr_delay` correspond to delays in minutes, with negative times representing an early departure/arrival  
- `air_time`, `distance` correspond to amount of time spent in air and distance between airports, in minutes and miles respectively

Dplyr has 6 key data transformation functions:  

- Pick observations by their values (`filter()`).  
- Reorder the rows (`arrange()`).  
- Pick variables by their names (`select()`).  
- Create new variables with functions of existing variables (`mutate()`).  
- Collapse many values down to a single summary (`summarise()`).  
- Change the scope of operation of functions to groups (`group_by()`).  

#### Exercises: Filter (5.2.4)
```{r eval=FALSE}
# Question 1
flights %>% filter(arr_delay >= 120)
flights %>% filter(dest == "IAH" | dest == "HOU")
flights %>% filter(carrier %in% c("UA", "AA", "DL"))
flights %>% filter(month %in% c(7, 8, 9))
flights %>% filter(dep_delay <= 0 & arr_delay > 120)
flights %>% filter(dep_delay >= 60 & dep_delay-arr_delay > 30)
flights %>% filter(0 <= hour & hour <= 6)

# Question 2
all.equal(flights %>% filter(hour %>% between(0,6)), flights %>% filter(0 <= hour & hour <= 6))

# Question 3
flights %>% filter(is.na(dep_time)) %>% nrow
flights %>% filter(is.na(dep_time)) %>% head
# Based on the other missing variables, these are likely canceled flights that never took off

# Question 4
NA ^ 0      # Any number to the 0th power is 1
NA | True   # Will always evaluate to True because only one of the sides needs to be True
False & NA  # Will always evaluate to False because only one of the sides needs to be False
# General rule: NA * any number except 0 is NA; If multiplied by 0, the product will always be 0
```

#### Exercises: Arrange (5.3.1)
```{r eval=FALSE}
# Question 1
flights %>% arrange(desc(is.na(dep_time))) %>% head

# Question 2
flights %>% arrange(desc(dep_delay)) %>% head
flights %>% arrange(dep_delay) %>% head

# Question 3
# There are quite a few ways to approach this question. One is by interpreting fastest to mean pure velocity (time in air/distance)
flights %>% arrange(desc(distance/air_time)) 
# Another way could be to see which flights made up the most time in the air, proportional to distance - this yields interesting results, and one explanation could be that the carrier is including consistent delays in its scheduled departure and arrival time
flights %>% arrange(desc((dep_delay-arr_delay)/distance))

# Question 4
# These questions can be tackled from two different perspectives as well -- length in terms of distance? Or length in terms of time?
flights %>% arrange(desc(distance)) # Longest by length
flights %>% arrange(air_time) # Shortest by time
```

Helper functions for use with `select`:  

- `starts_with("abc")`: matches names that begin with “abc”.
- `ends_with("xyz")`: matches names that end with “xyz”.
- `contains("ijk")`: matches names that contain “ijk”.
- `matches("(.)\\1")`: selects variables that match a regular expression. This one matches any variables that contain repeated characters.
- `num_range("x", 1:3)`: matches x1, x2 and x3.

`rename()` is a version of select that allows you to rename columns and keeps all other columns not explicitly mentioned as well

`mutate()` is useful for generating new columns and `everything()` can be useful when you want to move a column (or columns) to the beginning of the data frame like so:
```{r}
flights %>% head
flights %>% 
    mutate(air_time_made_up=dep_delay-arr_delay) %>% 
    select(air_time_made_up, everything()) %>% head
```

Forming groups (and summarizing them) are also useful, as we can group by each "date" and get the average departure delay per date, then sort to find the worst days to travel.
```{r}
flights %>% 
    group_by(year, month, day) %>% 
    summarise(delay=mean(dep_delay, na.rm=TRUE)) %>%
    arrange(desc(delay))
    
```

### Chapter 7 - Exploratory Data Analysis (EDA)
During EDA, we want to develop an understanding of the data to help us ask intriguing questions. 2 especially relevant questions are nearly always useful:  
1. What type of variation occurs within my variables?
2. What type of covariation occurs between my variables?  

During EDA, graphing is an effective way to begin understanding the variables present. Detecting outliers and understanding why they are outliers (and possibly logically inconsistent) is also important. One way outliers often manifest themselves, even if not necessarily visible, is with graph axes limits being stretched beyond what might be considered normal for the data that is shown.

We may want to understand variables in comparison to each other:
```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +     # 
  coord_flip()      # Flips the plot horizontally 
```

### Chapter 10 - Tibbles  
Tibbles are modernized `data.frame`s. Any data.frame can be converted into a tibble (though tibbles will never change input variable types/names and never creates row names) and tibbles can be created on their own like so:
```{r}
tibble(
  x = 1:5, 
  y = 1,                   # Single length inputs autopopulate to the appropriate length
  z = x ^ 2 + y
)

# Example with non-syntactic names
tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)

# Example transposed tibble
tribble(
  ~x, ~y, ~z,
  #--|--|----
  "a", 2, 3.6,
  "b", 1, 8.5
)
```
Tibbles are designed to print less and slightly more informatively, so they work well with big data.


### Chapter 11 - Data Import  

In general, using `data.table::fread()` or `readr::read_csv()` is much faster than base R's implementation of `read.csv()`. `read_csv` also automatically reads in dataframes as tibbles, has a progress bar, and is more reproducible than `read.csv`. 

Parse methods, such as `parse_logical()`, `parse_integer()`, and `parse_date()` are useful for converting strings or string vectors into their "true" data types.


### Chapter 12 - Tidy Data

Using tidy data makes analysis much simpler. Each datatset goes in a tibble and each variable goes in its own column.

`gather()` and `spread()` are two useful `tidyR` functions that can help transform non-tidy data (with variables in multiple columns or rows, respectively) into tidy data.`seperate()` is another function that helps if multiple "columns'" data is stuck within a single column (i.e.`123 / 3.1415`), and helps seperate that out into multiple columns. `unite()` is the exact opposite of `separate()` and can unite columns

### Chapter 13 - Relational Data

`semi_join(x, y)` keeps all observations in x that have a match in y
`anti_join(x, y)` drops all observations in x that have a match in y.

### Chapter 14 - Strings
